Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-04-12 SIR_manuscript.tex
updating the notations, clean up

\todo{the algebra needs to be changed based on JD's comment} The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns (\ali{is this clear or the "unkowns" needs to be defined? also * in sup for I and R?}. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
For brevity, we write $x\_u^* = S\_u^*/N_0$ and $x\_n^* = S\_n^*/N_0$.
We can then write $\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S$ for the \percap testing rate of infectious individuals near the disease-free equilibrium.

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
\Delta=& (A~ \theta\_w + B~ \theta\_c) / C\\
A=& \gamma \Big( \big(\gamma(\omega+\gamma) + \omega p_I F_I \big) x_n^* + (\omega+\gamma) F_I \Big)\\
B=& \omega p_I F_I (\gamma x_u^*+\omega) \\ 
C=& (\omega+\gamma) \Big(\gamma(\omega+\gamma)+F_I(\gamma+\omega p_I) \Big).
\end{aligned}
\end{equation}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
% \widetext
\begin{center}
\textbf{\large Appendix}
\end{center}
%%%%%%%%%% Merge with supplemental materials %%%%%%%%%%
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
% \setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{A\arabic{equation}}
\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\bibnumfmt}[1]{[A#1]}
\renewcommand{\citenumfont}[1]{A#1}
%%%%%%%%%% Prefix a "S" to all equations, figures, tables and reset the counter %%%%%%%%%%

@@

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are \ali{all $S\_u \& S\_n$ should be $S\_u^* \& S\_n^*$}
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_c)\,S\_u^*\\
S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_c)\,S\_n^* \\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right] \\
 =&\beta\left[ \begin {array}{c} x\_n^* \\ x\_u^* \\ 0 \\ 0 \end {array} \right]
        \left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right], \text{and}\\  V=&
 \left[ \begin {array}{cccc}  
F_I+\gamma&-\omega&0&0\\
-(1-p_I)F_I&\omega+\gamma&0&0\\
-p_I F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right].
\end{align}
The matrix inverse of $V$ is 
\begin{align}
\label{vinv}
V^{-1} =&
\frac{1}{\gamma C}
\left[ \begin {array}{cccc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)&0&0\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I)F_I & \gamma(\omega+\gamma)(F_I+\gamma)&0&0 \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I F_I & \gamma\omega p_I F_I & C \gamma/(\omega+\gamma) & 0 \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I F_I & \omega^2 p_I F_I & C \omega/(\omega+\gamma) & C
\end {array} \right],
\end{align}
where $C= \big( \gamma(\omega+\gamma)+(\gamma+\omega p_I)F_I \big) (\omega+\gamma)$. Note that all the columns of matrix $V^{-1}$ summ up to $C$.

The particular form of $F$ with two rows of zeros at the bottom, simplifies $G$ as the following block matrix.
\begin{equation}
\label{mat:G}
G = \left[ \begin {array}{cc}
G_{11}&G_{12}\\
0&0
\end {array} \right],
\end{equation}
where the 2 by 2 submatrix $G_{11}$ is 
\begin{equation}
\label{G11}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  x\_n^* \\ x\_u^*  \end {array} \right]
\left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right]
\left[ \begin {array}{cc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I)F_I & \gamma(\omega+\gamma)(F_I+\gamma) \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I F_I & \gamma\omega p_I F_I \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I F_I & \omega^2 p_I F_I
\end {array} \right].
\end{equation}

It is notable that matrix $F$ \eqref{FV} and consequently $G_{11}$ have rank one. That is $G_{11}$ has only one non-zero eigenvalue which we refer to it the basic reproduction number $\Rnum$ (defined as the spectral radius of $G$). Also the block matrix $G_{12}$ does not influence $\Rnum$ and all $S\_u$'s and $S\_n$'s are evaluated at the DFE \eqref{dfe} in matrix $F$ \eqref{FV} and the block matrix $G_{11}$ \eqref{mat:G}.

$\Rnum$ has a complicated expression and depend on the type of analysis, the expression can be simplifed and presented accordingly. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $A$, $B$ and $C$, specified in \eqref{eq:abc}. This is intended to show that $\Rnum$ decreases as the strength of isolation, shown by $\theta\_w$ and $\theta\_c$, increases.

It remains hard to show the reduction of $\Rnum$ with respect to \percap testing intensity, $\rho$, and the speed of the test return, $\omega$, for the feasible ranges of the parameters, i.e., $\omega>0$, $0 \leq \rho<\omega$, $0 \leq \theta\_w \leq \theta\_c \leq 1$ and $\frac{w_I}{w_S}\geq 1$. One easy way to do such an analysis is based on the fact that $\rho$ is \percap rate so for a large population size, $N_0 \gg 1$, $\rho$ is very small or close to 0, comparing to $N_0$. This provides a base to linearly approximate $\Rnum$ when $\rho$ is close to 0, and use this approximation to analyze the behaviour of $\Rnum$ with respect to $\omega$ (see section \ref{taylor}). In the next section we provide an equivalent representation of $\Rnum$ for our model to provide to follow proving that more testing intensity decreases $\Rnum$ for a general rane of parameters.

@@

This is to prove that $\pro \Delta > 0$ with $\Delta$ defined in Eq. \eqref{eq:abc}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eqs. \eqref{R0}.

Another representation of matrix $G_{11}$ in \eqref{G11} is as follows.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  x\_n^* \\ x\_u^*  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_1-C_2\end {array} \right],
\end{equation}
where $C$ is in Eq. \eqref{eq:abc}, and $C_1$ and $C_2$ are 
\begin{align}
\label{eq:C12}
C =& (\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,F_I)\\
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) F_I,\\
C_2 =& (\omega+\gamma)\gamma^2\,\theta\_w+(\theta\_w-\theta\_c)\gamma\,p_I\,F_I.
\end{align}
$\Rnum$ is in the same form as in Eq. \eqref{R0}  $$\Rnum= \frac{\beta}{\gamma} (1-\Delta),$$
with equivalent but simplified $\Delta$ as
\begin{equation}
\label{eq:del2}
\Delta= \frac{C_1+C_2\,x\_n^*}{C}.
\end{equation}
The behavior of $\Rnum$ with respect to $\rho$ is determine by $\pro \Delta$
\begin{align}
\label{eq:dd2dr}
\pro \Delta=& \frac{1}{d} (a\,\rho^2-b\,\rho+c),
\end{align}
where
\begin{align}
\label{eq:abcd}
a=& ((\omega+\gamma)\gamma -\omega \frac{w_I}{w_S} (\omega\,p_I+\gamma)) D_1,\\
b=& 2\gamma\omega(\omega+\gamma) D_1,\\
c=& \omega^2 (\omega+\gamma)^2 D_2, \\
d=& \frac{\omega}{\gamma}(\omega+\gamma)\Big( \big( \gamma(\omega+\gamma)-\omega \frac{w_I}{w_S}(\omega\,p_I+\gamma)\big) \rho-\omega\gamma(\omega+\gamma) \Big)^2,
\end{align}
and $D_1$ and $D_2$ as 
\begin{align}
\label{eq:D1D2}
D_1=& (\omega+\gamma)\gamma\theta\_w+(\theta\_c-\theta\_w)\omega^2\,p_I\,\frac{w_I}{w_S}, \\
D_2=& (\omega\,\frac{w_I}{w_S}+\gamma)\gamma\theta\_w+\omega^2\,p_I\,\frac{w_I}{w_S}\,\theta\_c.
\end{align}
Note the following conditionbs and assumptions of the parameters of the model;
\begin{align}
\label{cond}
& \omega>0, \\
& 0 \leq \rho<\omega,\\ 
& 0 \leq \theta\_w \leq \theta\_c \leq 1, \\
& \frac{w_I}{w_S}\geq 1.
\end{align}
Given the parameter restrictions, we would like to determine the sign pattern of $\pro \Delta$ given the quadratic form of it in Eq. \eqref{eq:dd2dr}. Thus, it is important to know the sign of the coefficients $a$, $b$ and $c$ of the polynomial in Eq. \eqref{eq:dd2dr}. Note that with scaler $d>0$, it has no effect on the sign of $\pro \Delta$. Since $D_1\geq 0$ and $D_2\geq 0$, $b\geq 0$ and $c\geq 0$, sign of $a$ depends on the interplay between the involved parameters. Note that $D_1 \leq D_2$. (\todo{still working on}).

>>>>

\todo{the algebra needs to be changed based on JD's comment} The Disease-Free Equilibrium (DFE) for the SIR model, Eqs. \eqref{model}, is given by setting the differential equations to 0 and solving for the unknowns (\ali{is this clear or the "unkowns" needs to be defined? also * in sup for I and R?}. The DFE is
\begin{equation}
\label{dfe}
S\_n^*= \frac{\rho}{\omega} N_0, \ S\_u^*= N_0-S\_n^*, \text{~and~} I_j=R_j=0 \ \text{for all }j.
\end{equation}
For brevity, we write $x\_u^* = S\_u^*/N_0$ and $x\_n^* = S\_n^*/N_0$.
We can then write 
\begin{equation}
\label{eq:fi}
\hat F_I = (\omega\rho/(\omega-\rho))w_I/w_S
\end{equation}
for the \percap testing rate of infectious individuals near the disease-free equilibrium.

The basic reproduction number, $\Rnum$, was calculated by using the next generation matrix method developed by \cite{van2002reproduction}. $\Rnum$ is
\begin{equation}
\label{R0}
\Rnum= \frac{\beta}{\gamma} (1-\Delta), 
\end{equation}
where
\begin{equation}
\begin{aligned}
\label{eq:abc}
\Delta=& (A~ \theta\_w + B~ \theta\_c) / C\\
A=& \gamma \Big( \big(\gamma(\omega+\gamma) + \omega p_I \hat F_I \big) x_n^* + (\omega+\gamma) \hat F_I \Big)\\
B=& \omega p_I(\gamma x_u^*+\omega)\hat F_I \\ 
C=& (\omega+\gamma) \Big(\gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I\Big).
\end{aligned}
\end{equation}

@@

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage
% \widetext
\begin{center}
\textbf{\large Appendix}
\end{center}
\setcounter{equation}{0}
\setcounter{figure}{0}
\setcounter{table}{0}
% \setcounter{page}{1}
\makeatletter
\renewcommand{\theequation}{A\arabic{equation}}
\renewcommand{\thefigure}{A\arabic{figure}}
\renewcommand{\bibnumfmt}[1]{[A#1]}
\renewcommand{\citenumfont}[1]{A#1}

@@

where parameters are specified in Table \ref{tab:params}. The next generation matrix for this model is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. 
Matrices $F$ and $V$ are
\begin{align}
\label{FV}
F =& \beta/N_0 \left[ \begin {array}{cccc} 
S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_w)\,S\_u^* & (1-\theta\_c)\,S\_u^*\\
S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_w)\,S\_n^* & (1-\theta\_c)\,S\_n^* \\ 
0&0&0&0\\
0&0&0&0
 \end {array} \right] \\
 =&\beta/N_0 \left[ \begin {array}{c} S\_n^* \\ S\_u^* \\ 0 \\ 0 \end {array} \right]
        \left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right], \text{and}\\  V=&
 \left[ \begin {array}{cccc}  
\hat F_I+\gamma&-\omega&0&0\\
-(1-p_I)\hat F_I&\omega+\gamma&0&0\\
-p_I \hat F_I&0&\omega+\gamma&0\\
0&0&-\omega&\gamma
\end {array} \right].
\end{align}
The matrix inverse of $V$ is 
\begin{align}
\label{vinv}
V^{-1} =&
\frac{1}{\gamma C}
\left[ \begin {array}{cccc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)&0&0\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I) \hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma)&0&0 \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I & C \gamma/(\omega+\gamma) & 0 \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I & C \omega/(\omega+\gamma) & C
\end {array} \right],
\end{align}
where $C= \big( \gamma(\omega+\gamma)+(\gamma+\omega p_I)\hat F_I \big) (\omega+\gamma)$ and $\hat F_I$ is the \percap testing rate for the infected people and represented in Eq. \eqref{eq:fi}. Note that all the columns of matrix $V^{-1}$ summ up to $1/\gamma$.

The particular form of $F$ with two rows of zeros at the bottom results in the following blocked form of matrix $G$.
\begin{equation}
\label{mat:G}
G = \left[ \begin {array}{cc}
G_{11}&G_{12}\\
0&0
\end {array} \right],
\end{equation}
where both blocked matricies $G_{11}$ and $G_{12}$ are 2 by 2. Given the upper triangular form of matrix $G$, the basic reproduction number $\Rnum$ (defined as the spectral radius of matrix $G$) is only determined by the blocked matrix $G_{11}$,
\begin{equation}
\label{G11}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c} (\omega-\rho)/\omega \\ \rho/\omega \end {array} \right]
\left[ \begin {array}{cccc} 1, \, 1-\theta\_w, \, 1-\theta\_w, \, 1-\theta\_c \end {array} \right]
\left[ \begin {array}{cc}
\gamma(\omega+\gamma)^2 & \gamma\omega(\omega+\gamma)\\ \noalign{\medskip}
\gamma(\omega+\gamma)(1-p_I)\hat F_I & \gamma(\omega+\gamma)(\hat F_I+\gamma) \\ \noalign{\medskip}
\gamma(\omega+\gamma) p_I \hat F_I & \gamma\omega p_I \hat F_I \\ \noalign{\medskip}
\omega(\omega+\gamma) p_I \hat F_I & \omega^2 p_I \hat F_I
\end {array} \right].
\end{equation}
It is notable that matrix $F$ \eqref{FV} has rank one and consequently $G_{11}$ does so. That is $G_{11}$ has only one non-zero eigenvalue which is $\Rnum$.

The expression of $\Rnum$ has a complicated form with all of the model parameters involved. This expression can be simplifed and represented given the specific form of matrix $G_{11}$ \eqref{G11}. For the purpose of simplicity we present $\Rnum$ in the manuscript in terms of expressions $A$, $B$ and $C$, specified in \eqref{eq:abc}. This is intended to show that $\Rnum$ decreases as the strength of isolation, shown by $\theta\_w$ and $\theta\_c$, increases.

It remains hard to show that the reproduction number $\Rnum$ is decreasing with respect to \percap testing intensity, $\rho$, and the speed of the test return, $\omega$, for the feasible ranges of the parameters, that is
\begin{align}
\label{cond}
& \omega>0, \\
& 0 \leq \rho<\omega,\\ 
& 0 \leq \theta\_w \leq \theta\_c \leq 1, \\
& \frac{w_I}{w_S}\geq 1.
\end{align}
One way to do such an analysis is based on the fact that $\rho$ is \percap rate so for a large population size, $N_0 \gg 1$, $\rho$ is very small or close to 0, comparing to $N_0$. This provides a base to linearly approximate $\Rnum$ when $\rho$ is close to 0, and use this approximation to analyze the behaviour of $\Rnum$ with respect to $\omega$ (see section \ref{taylor}). 
In the next section we provide an equivalent representation of $\Rnum$ to provide a ground to prove that more testing intensity decreases $\Rnum$ for a general rane of parameters.

@@

This is to prove that $\pro \Delta > 0$ with $\Delta$ defined in Eq. \eqref{eq:abc}, and thus $\pro \Rnum < 0$, where $\Rnum$ is the basic reproduction number, given in Eqs. \eqref{R0}. We can rewrite matrix $G_{11}$ in \eqref{G11} in the following form to simplify the calculations.
\begin{equation}
\label{G112}
G_{11} = \frac{\beta}{\gamma \, C} 
\left[ \begin {array}{c}  (\omega-\rho)/\omega \\ \rho/\omega  \end {array} \right]
\left[ \begin {array}{cccc} 
C-C_1, C-C_1-C_2\end {array} \right],
\end{equation}
where $C$ is the same as the one in Eq. \eqref{eq:abc}, i.e.,
$$C=(\omega+\gamma)(\gamma(\omega+\gamma)+(\omega\,p_I+\gamma)\,\hat F_I),$$
and $C_1$ and $C_2$ are 
\begin{align}
\label{eq:C12}
C_1 =& (\omega+\gamma)(\theta\_w \, \gamma+\theta\_c\, \omega\, p_I) \hat F_I,\\
C_2 =& (\omega+\gamma)\gamma^2\,\theta\_w+(\theta\_w-\theta\_c)\gamma\,p_I\,\hat F_I,
\end{align}
where $\hat F_I$ is given in Eq. \eqref{eq:fi}.
$\Rnum$ is in the same form as in Eq. \eqref{R0}  
$$\Rnum= \frac{\beta}{\gamma} (1-\Delta),$$
with equivalent but simplified $\Delta$ in Eq. \eqref{eq:abc} as
\begin{equation}
\label{eq:del2}
\Delta= \frac{C_1+C_2\,\rho/\omega}{C}.
\end{equation}
The behavior of $\Rnum$ with respect to $\rho$ is determine by $\pro \Delta$
\begin{align}
\label{eq:dd2dr}
\pro \Delta=& \frac{1}{d} (a\,\rho^2-b\,\rho+c),
\end{align}
where
\begin{align}
\label{eq:abcd}
a=& ((\omega+\gamma)\gamma -\omega \frac{w_I}{w_S} (\omega\,p_I+\gamma)) D_1,\\
b=& 2\gamma\omega(\omega+\gamma) D_1,\\
c=& \omega^2 (\omega+\gamma)^2 D_2, \\
d=& \frac{\omega}{\gamma}(\omega+\gamma)\Big( \big( \gamma(\omega+\gamma)-\omega \frac{w_I}{w_S}(\omega\,p_I+\gamma)\big) \rho-\omega\gamma(\omega+\gamma) \Big)^2,
\end{align}
and $D_1$ and $D_2$ as 
\begin{align}
\label{eq:D1D2}
D_1=& (\omega+\gamma)\gamma\theta\_w+(\theta\_c-\theta\_w)\omega^2\,p_I\,\frac{w_I}{w_S}, \\
D_2=& (\omega\,\frac{w_I}{w_S}+\gamma)\gamma\theta\_w+\omega^2\,p_I\,\frac{w_I}{w_S}\,\theta\_c.
\end{align}

Given the parameter restrictions \ref{cond}, we would like to determine the sign pattern of $\pro \Delta$ given the quadratic form of it in Eq. \eqref{eq:dd2dr}. Thus, it is important to know the sign of the coefficients $a$, $b$ and $c$ of the polynomial in Eq. \eqref{eq:dd2dr}. Note that with scaler $d>0$, it has no effect on the sign of $\pro \Delta$. Since $D_1\geq 0$ and $D_2\geq 0$, $b\geq 0$ and $c\geq 0$, sign of $a$ depends on the interplay between the involved parameters. Note that $D_1 \leq D_2$.
